# Introduction

- write section of progress to date
- animals who have their needs met in a given area minimize their movement to avoid wasting energy. movement to a new area is often caused by inferior resource availability

## Objectives

This project will produce a global raster of a new environmental stochasticity index, and new quantitative methods for animal movement. Findings will inform design of protected areas and assist in conserving Canada's wildlife, particularly in light of Canada's pledge to protect 30% of its landmass and oceans by 2030[@ryan_patrick_jones_pm_2021], although only 13% of Canada’s area is currently protected[@government_of_canada_canadian_2021]. Local Indigenous groups will be included in the research process, in the hope of forming long-lasting co-operative relationships.

I will use simulation studies and an unprecedented and conservation-relevant animal tracking dataset (>1500 animals, 77 globally-distributed species) to provide the most detailed investigation into how animal spatial needs change with environmental stochasticity to date. This work has four key objectives: (i) estimating individuals' spatial requirements in a way which is insensitive to variation in sampling protocols and data quality; (ii) quantifying environmental stochasticity and its effects; (iii) estimating between-species trends using models that are robust to commonly-found issues (e.g., correlations within species); and (iv) understanding how Traditional Indigenous Knowledge can be integrated into large-scale ecological research and conservation planning within a framework that acknowledges both Traditional Indigenous Knowledge and Western science[@kutz_two-eyed_2019].

To achieve these aims: (i) I will use continuous-time models [@johnson_continuous-time_2008] that do not depend on sampling frequency. Such models will allow me to use the entirety of the data rather than aggregated data (e.g., daily averages), as aggregated data contain less information on sample variance and can lead to biased estimates. (ii) I will produce a global, time-varying raster of environmental stochasticity which accounts for productivity[@nilsen_can_2005], weather, and climate (e.g., precipitation, temperature), as well as the frequency of extreme events (e.g., heat domes, flooding, forest fires). I will then use the raster to estimate the effect of stochasticity on home range sizes. (iii) I will use a hierarchical approach[@pedersen_hierarchical_2019] to estimate common trends and variances within and between populations, species, and data collection methods. (iv) I will collaborate closely with with various Indigenous groups and include any Traditional Knowledge and practices they wish to include in my project. The research will be published in open-access journals and all code will be publicly available.

## Environmental stochasticity

- Chaotic Markov chains?
- https://www.greensensefarms.com/home
- https://onlinelibrary.wiley.com/doi/10.1111/plb.12085
- https://spinoff.nasa.gov/Spinoff2018/cg_7.html
- predictable events: lizard's alternating gate to minimize contact time on hot sand
- The $\gamma$-variance between spatiotemporal points at a given (spatiotemporal, multidimensional) lag $\vec h$ is calculated as $$\gamma(\vec h) = \frac 12 \mathbb E\left[Z(\vec x + \vec h) - Z(\vec x) \right]^2 = \frac 12 Var\left[Z(\vec x + \vec h) - Z(\vec x) \right]$$ [@bachmaier_variogram_2011]. The $\frac 12$ is because we don't need to repeat each pair (e.g., $V((a, b), (b, a)) = 2 V((a, b))$). View $\gamma(\vec h)$ as a function of $\vec h$ rather than a fixed value.

Different types of environmental stochasticity:

* spatial stochasticity:
  - environmental patches and niches,
  - habitat heterogeneity
* temporal stochasticity:
  - events (heat domes, fires, tornadoes, storms)
  - events are hard to predict when:
    - they do not occur frequently within an organism's generation time
    - they occur frequently but unpredictably
common but stochastic (i.e., $p \approx 0.5$) 
* spatio-temporal stochasticity (in the inclusive sense, not just the interaction):
  - alternate stable states
  - ephemeral habitats following stochastic events (e.g. burned forest, flooded area -- but compare to yearly flooded areas such as temperate forests)

**Small-scale patch structure in North American and South**:
* Alone, fire homogenizes while herbivory diversifies, but together they diversify the most (Fuhlendorf and Engle 2004; Archibald
et al. 2005; Collins and Smith 2006)

### The time scale of stochastic events

When quantifying stochastic events, the severity and frequency of the events should be scaled relative to an organism's generation time [@ref]. Events which are stochastic but occur frequently will lead an organism to develop protective mechanisms, particularly if such events pose a high risk of mortality or injury. *How do animals from areas with frequent extreme events (e.g., sand storms) avoid damage? Do they use dens or develop a resistance?* Bison (*Bison bison*) have fur to protect them from cold winds, and they stand in groups facing the wind to decrease heat loss and damage from the cold.

As the occurrence of an event become more unpredictable (i.e., less common), the effect of such events may be very destructive in the moment, but recovery may be possible. In contrast, if the magnitude of an event and its damage become more stochastic, recovery may be harder to achieve. As extreme events become more frequent, the variance in occurrence increases until the events become common. For a Bernoulli random variable with probability of occurrence $p$, $Y \sim Ber(p)$, the variance (i.e., stochasticity) is maximized when $p = 0.5$ and lowest when the event occurs almost never ($p \approx 0$) or or almost always ($p \approx 1$).

As a extreme events becomes more common, organisms may begin to adapt to their occurrence, as long as the change in frequency is sufficiently gradual to allow adaptation.

If stochastic events become common, animals may learn to "store for a rainy day". For instance, if the probability of a downpour at night in a desert is high but highly unpredictable ($p \approx 0.5$), animals and plants may develop protective mechanisms [e.g. cacti closing their flowers at sunset; @stanton_light_2010]. When extreme events occur frequently, . If a dangerous event becomes more frequent and unpredictable, organisms may tend to take advantage of favorable times so they can resist more adverse times. Frequent destructive events may be less dangerous than infrequent destructive events. Events which do not occur frequently during an organism's generation time or lifespan (e.g., 2 times or less) may not be considered as threats worth preparing for. In contrast, stochastic and destructive events which happen frequently become part of an organism's habitat, and such organisms may adapt to their occurrence and even learn to predict them [@ref].

The time to re-colonization of a habitat [also see the *rescue effect* sensu @ref] will not necessarily be proportional to the organisms' generation times. The movement of organisms with short generation times tends to be shorter because generation time correlates with mass [@ref]. Thus, organisms with short generation times [e.g., aquatic invertebrates, see @kolasa_rapoports_1998] are more likely to move slowly or be moved by large and potentially unpredictable events (e.g., waves, tides, sea currents, storms). 

```{r, echo = FALSE, eval = FALSE}
library('ggplot2')
library('tibble')
theme_set(theme_bw())
x <- tibble(p = seq(1e-4, 1 - 1e-4, by = 1e-4),
            var = 1 * p * (1 - p)) # a single Ber(p) event
ggplot(x) +
  geom_line(aes(p, var)) +
  labs(x = 'Probabilty of the event occuring (p)',
       y = expression(Event~variance~(sigma^2~'='~1%*%p%*%(1-p))))
```

```{r, echo = FALSE, eval = TRUE, message = FALSE, fig.cap= "Realizations of Bernoulli processes for different probability parameters, $p$. At low values of $p$, successes occur unpredictably", fig.align='center'}
library('ggplot2')
library('dplyr')
library('tidyr')
theme_set(theme_bw())
expand_grid(t = 1:50,
            p = seq(0, 1, by = 0.1)) %>%
  mutate(var = 1 * p * (1 - p),
         y = rbinom(n = n(), size = 1, prob = p)) %>%
  filter(y == 1) %>% # remove failures
  ggplot() +
  geom_hline(yintercept = seq(0, 1, by = 0.1), alpha = 0.1) +
  geom_point(aes(t, p), shape = 1, size = 2) +
  geom_point(aes(t, p, color = var)) +
  scale_x_continuous('Time', breaks = NULL) +
  scale_y_continuous('Probability of occurrence (p)', limits = c(0, 1)) +
  scale_color_gradient(expression(sigma^2~'='~p(1-p)), low = 'white', high = 'black') +
  theme(panel.grid = element_blank())
```

In the case of obligate symbiosis or interaction, a species' ability to change its range or adapt will likely depend strongly on its associate's ability to move or adapt, too.

Large-scale studies such as this one are useful to detect general and common trends, while small scale studies such as the work of @kolasa_rapoports_1998 can provide more information on the small-scale drivers of phenomena.

An animal's ability to alter its behavior and movement in response to environmental conditions is essential in stochastic or changing landscapes. Whether the changes be due to highly variable but (potentially) predictable changes such as climate change and the weather, or whether they be due to more stochastic events (e.g., fires, floods, oil spills), an animal's ability to adapt increases its odds of survival. @rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on variable environmental events and cues such as available forage biomass, hunting pressure, snow fall, and snow melt. Birds have also shown to change their their migration as the climate changes, including trans-Saharian migrants [@jonzen_rapid_2006].

An organism's dependence on stochastic cycles such as changes in temperature, precipitation, and resource availability is more efficient and more plastic than depending on purely deterministic cycles (e.g., photoperiod). While this thesis focuses on the movement and spatial use of animals, the adaptability (or lack thereof) of non-animal organisms is also crucial. The ability of most animals to rely on visual cues and move accordingly greatly increases their adaptability and plasticity, particularly for those animals that are able to move large distances over short periods of time (e.g., birds, large vertebrates). Since plants and fungi cannot rely on visual cues nor move as easily, they must depend more on effective reproduction timing, seed dispersal, and evolution of the species rather than the fitness of the individual [@ref?]. Recent changes in flowering and leafing times of plants have been estimated, but changes are occurring faster than predicted by experimental estimates by as much as 4.0-8.5 times [@wolkovich_warming_2012]. This discrepancy between experimental and observational estimates indicates the importance of interpreting experimental results carefully and conditionally on the model and data. The underestimation suggests experimental models fail to account for important factors, such as changes in wind disturbance and soil moisture, shifts in genotypes, and community structure, as well as loss of species diversity [@ref on the change in phenology with species loss].

See *The effects of phenological mismatches on demography*.

@rickbeil_plasticity_2019:
* Many species use seasonal migration to track emerging resources (Fryxell, Greever, & Sinclair, 1988), reduce predation risk (Hebblewhite & Merrill, 2009), or avoid disease and parasites (Altizer, Bartel, & Han, 2011).
* Migrations to track resources are often timed to coincide with peak plant and insect emergence (Aikens et al., 2017; Jonzén et al., 2006; Merkle et al., 2016)
* there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006; Lameris et al., 2018) and highlighting a potential measure of resilience in the face of climate change
* Elk departer earlier from highly-productive areas? This goes against the idea that animals move as a consequence of need. This could be a spurious correlation or maybe a correlation to an unaccounted effect?
* Elk left early when exposed to hunting more
* 
