\newpage

# Project timeline {#timeline}

```{r, echo = FALSE, fig.cap='Visual representation of the estimated timeline for my Master\'s project.', fig.width=14, fig.height=8, message=FALSE, out.width='100%'}
library('tibble')    # for fancy data frames
library('dplyr')     # for piping and data wrangling
library('lubridate') # for working with dates
library('stringi')   # for working with character strings
library('ggplot2')   # for plotting the timeline
# library('ggrepel')   # for non-overlapping ggplot labels
theme_set(theme_classic() +
            theme(axis.line.y = element_blank(),
                  legend.position = 'none'))

# curstom color-vision deficient palette
pal <- c('#4477AA', '#ff8c00', '#66CCEE', '#009900',
                  '#CCBB44', '#AA3377', '#EE6677', '#BBBBBB')
                  
t_breaks <- tibble(dates = c('2021-09-01',
                             '2022-01-01', '2022-05-01', '2022-09-01',
                             '2023-01-01', '2023-05-01', '2023-09-01',
                             '2024-01-01', '2024-05-01', '2024-09-01',
                             '2025-01-01', '2025-05-01', '2023-09-01') %>%
                     date(),
                   dec = decimal_date(dates),
                   strings = paste(month.abb[month(dates)], year(dates)))

tasks <-
  readr::read_csv('timeline.csv', col_types = 'ccDD') %>%
  mutate(header = purrr::map2_lgl(section, task, grepl),
         section = factor(section,
                          levels = c('Other', 'Chapter 5', 'Chapter 4',
                                     'Chapter 3', 'Chapter 2', 'Chapter 1',
                                     'Preparatory', 'Coursework'))) %>%
  arrange(section, desc(start), desc(end)) %>%
  mutate(dec_start = lubridate::decimal_date(start),
         dec_end = lubridate::decimal_date(end),
         dec_mid = purrr::map2_dbl(dec_start, dec_end, \(x, y) (x + y) / 2),
         task = factor(task, levels = unique(task)),
         id = 1:n(),
         completed = end <= Sys.Date(),
         # task = stri_replace_all(task, regex = ':', replacement = ':\n'),
         # re-format all special characters (only needed if parse = TRUE)
         # task = stri_replace_all(task, regex = ' ', replacement = '~'),
         # task = stri_replace_all(task, regex = '\'', replacement = "\\'"),
         # task = stri_replace_all(task, regex = '~in~', replacement='~\'in\'~'),
         # task = stri_replace_all(task, regex = ',', replacement = ''),
         # task = stri_replace_all(task, regex ='for~', replacement='\'for\'~'),
         # make headers bold (removed because numbers cannot be bold)
         #task = if_else(header, paste0('bold(', task, ')'), task),
  )

# tasks %>%
#   transmute(Task = task, `Start date` = start, `End date` = end) %>%
#   knitr::kable(caption = 'Estimated timeline for my Master\'s project.')
ggplot(tasks, aes(y = id)) +
  geom_hline(aes(yintercept = id + 0.5), filter(tasks, header), alpha = 0.15) +
  geom_vline(xintercept = decimal_date(Sys.Date()), lty = 'dashed', alpha = 0.5) +
  geom_errorbar(aes(xmin = dec_start, xmax = dec_end, color = section),
                width = 0, linewidth = 2, alpha = 0.75) +
  #' use `ggrepel::geom_label_repel()` for repelling labels
  geom_text(aes(2020.2, label = task, color = header), hjust = 'left',
            size = 3, parse = FALSE) +
  scale_x_continuous(NULL, breaks = t_breaks$dec, labels = t_breaks$strings,
                     expand = c(0, 0)) +
  scale_y_continuous(NULL, breaks = NULL) +
  scale_color_manual(NULL, values = c('black', 'grey40', pal),
                     breaks = c(TRUE, FALSE, levels(tasks$section)))
```

## Progress to date

```{r, eval = FALSE}
filter(tasks, completed) %>%
  pull(task) %>%
  as.character() %>%
  cat(sep = '\n')
```

- Created and organized the GitHub repository  ([https://github.com/StefanoMezzini/hr-environ-stoch-masters](https://github.com/StefanoMezzini/hr-environ-stoch-masters));
- Completed all coursework necessary for a Master's degree;
- Reviewed literature and have begun making theoretical predictions;
- Living Data Project Working group: calculated new Canadian Living Planet Index (cLPI);
- Data analyst: estimated effects of human activity on tapir movement using `ctmm` and `mgcv`. The manuscript has been accepted and is in press (preprint is available at [https://www.biorxiv.org/content/10.1101/2021.11.12.468362v1](https://www.biorxiv.org/content/10.1101/2021.11.12.468362v1), and the code and data are available at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs);
- Data analyst: estimated changes paleolimnological time series from endorheic lakes due to 8-m lake-level variation using Hierarchical Generalized Additive Models (HGAMs) and location-scale HGAMs fit with `mgcv`. The manuscript is available at [https://doi.org/10.1002/lno.12054](https://doi.org/10.1002/lno.12054), while code and data are available at [https://github.com/simpson-lab/kenosee-white-bear](https://github.com/simpson-lab/kenosee-white-bear);
- Presented the background theory behind the thesis and some simulations as a guest lecturer in Dr. Noonan's BIOL 417 course (Evolutionary Ecology).
